Well-quasi-orderings on word languages

### 2025-09-30 src/conclusion.tex
new: add refs + morhpic words + todos
new: morphic sequences factor complexity + moving stuff around

\paragraph*{Graph Classes}
\todo[inline]{aliaume: retalk about \cite{ALM17} and \cite{L24:arxiv:v2}}

>>>>

\subsubsection{Factor Complexity}

\AP Let us conclude this section with a few remarks on the notion of \kl{factor
complexity} of languages. Recall that the \intro{factor complexity} of a
language $L \subseteq \Sigma^*$ is the function $f_L: \Nat \to \Nat$ such that
$f_L(n)$ is the number of distinct words of size $n$ in $L$. A language $L$ has
\intro{exponential factor complexity} if there exists a constant $C > 1$ such
that $f_L(n) \geq C^n$ for infinitely many $n \in \Nat$. A language $L$ has
\intro{subaffine factor complexity} if $f_L(n) \leq C \cdot n + K$ for some
constants $C, K > 0$ and all $n \in \Nat$.
We extend the notion of \kl{factor complexity} to infinite and bi-infinite words
as the \kl{factor complexity} of their set of finite \kl{infixes}.

While there clearly are languages with \kl{subaffine factor complexity} that are
not \kl{well-quasi-ordered} for the \kl{infix relation}, such as 
the language $L \defined \dwset{ a b^* a }$;
one would expect that languages that are \kl{well-quasi-ordered} for the
\kl{infix relation} would have a low \kl{factor complexity}.

\AP It is the case that \kl{bounded languages}, studied in
\cref{infixes-bounded:sec}, have a polynomial factor complexity: such a
language is by definition included in $w_1^* \cdots w_n^*$ for some words $w_1,
\dots, w_n$, and an infix is a choice of the number of occurrences of each
$w_i$ in the infix, plus a choice of beginning and ending of the infix in the
leftmost and rightmost $w_i$. We showed in \cref{bounded-language:thm} and
particularly in \cref{bounded-language:lem}, that \kl{bounded languages} that
are \kl{well-quasi-ordered} for the \kl{infix relation} are in fact included in
some finite union of languages of the form $w_1^* w_2 w_3^*$, hence have
actually a quadratic factor complexity. However,
\cref{exponential-factor-complexity:lem} shows that there are
\kl{downwards-closed} languages that are \kl{well-quasi-ordered} for the
\kl{infix relation} but have an \kl{exponential factor complexity}.

\begin{lemma}
  \label{exponential-factor-complexity:lem}
  There exists a language $L$ with \kl{exponential factor complexity}
  that is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}
\begin{proof}
  We leverage results from \cite{CAKA97} on Toeplitz words.
  Namely, the 
  $(5, 3)$-Toeplitz word is \kl{uniformly recurrent} 
  \cite[p. 499]{CAKA97},
  but has an exponential factor complexity 
  \cite[Theorem 5]{CAKA97}.
\end{proof}

Our understanding of the situation is that whenever a computational model is
fixed to represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation}, the \kl{factor complexity} of the represented languages drops to a
quadratic one. This is the case for languages recognized by \kl{amalgamation
systems} \cref{amalgamation-systems:subsec}, that are \kl{bounded languages}
when they are \kl{well-quasi-ordered} (\cref{infix-amalgamation:thm}). 
This is
also the case for languages described as the infixes of a finite set of pairs
of \kl{morphic sequences}. Indeed, the factor complexity of a \kl{morphic
sequence} that is \kl{uniformly recurrent} is linear \cite[Theorem 24]{NIPR09}.
and therefore the factor complexity of the set of infixes of a bi-infinite word
obtained as $\bar{w_1} u w_2$ where $w_1, w_2$ are \kl{morphic sequences} and
$u$ is a finite word is quadratic.

### 2025-09-30 src/infixes.tex
new: morphic sequences factor complexity + moving stuff around

Let us first remark that if $S$ is a \kl{chain} for the \kl{suffix relation}
and $P$ is a \kl{chain} for the \kl{prefix relation}, then $SP$ is
\kl{well-quasi-ordered} for the \kl{infix relation}. This proves the (easy)
right-to-left implication of \cref{bounded-language:thm}. Furthermore, any
language $L$ included in a finite union of products $S_i \cdot P_i$ where $S_i$
is in fact a \kl{bounded language}.

@@

    Now, $\oWidth{L} \leq \sum_{i = 1}^n \oWidth{L_i} < 3 \cdot \omegaOrd <
    \omegaOrd^2$. We conclude that the \kl{ordinal width} of $L$ is less than
    $\omegaOrd^2$. Finally, the inequality $\oType{L} \leq \oWidth{L} \oComProd
    \oHeight{L} < \omega \oComProd \omega^2 = \omegaOrd^3$ allows us to conclude.
\end{proofof}

\AP
Let us conclude this section with a few remarks on the notion of \kl{factor
complexity} of languages. Recall that the \intro{factor complexity} of a
language $L \subseteq \Sigma^*$ is the function $f_L: \Nat \to \Nat$ such that
$f_L(n)$ is the number of distinct words of size $n$ in $L$. A language $L$ has
\intro{exponential factor complexity} if there exists a constant $C > 1$ such
that $f_L(n) \geq C^n$ for infinitely many $n \in \Nat$.

\begin{lemma}
  \label{exponential-factor-complexity:lem}
  There exists a language $L$ with \kl{exponential factor complexity}
  that is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}
\begin{proof}
  We leverage results from \cite{CAKA97} on Toeplitz words.
  Namely, the 
  $(5, 3)$-Toeplitz word is \kl{uniformly recurrent} 
  \cite[p. 499]{CAKA97},
  but has an exponential factor complexity 
  \cite[Theorem 5]{CAKA97}.
\end{proof}

It is the case that \kl{bounded languages} have at most a polynomial factor
complexity, and that the \kl{Thue-Morse sequence} has a linear factor
complexity. More generally, as we will see, most ``concrete'' examples of
infinite words have a low factor complexity: \kl{automatic sequences} have at
most a linear factor complexity, and \kl(morphic){morphic sequences} have at
most a polynomial factor complexity. In that sense, the
\cref{exponential-factor-complexity:lem} shows that one can still construct
``pathological'' examples of infinite words that are \kl{well-quasi-ordered}
for the \kl{infix relation}, but have a high factor complexity.

\subsection{Decision Procedures}

\AP As we have demonstrated, infinite (or bi-infinite words) can be used to
represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation} by considering their set of finite \kl{infixes}. We can leverage the
theory of \kl{automatic sequences}, that is tailored to study the regularity of
infinite words to build a decision procedure for \kl{well-quasi-ordered}
languages for the \kl{infix relation}. A sequence $w \in \Sigma^\Nat$ is
\intro(sequence){automatic} whenever there exists a base $b \in \Nat$ and a
finite automaton that can compute the $i$-th letter of $w$ given as input the
number $i$ in base $b$. The \kl{Thue-Morse sequence} is an example of an
\kl{automatic sequence}.

\begin{theorem}
    \label{automatic-wqo:thm}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}.
\end{theorem}
\begin{proof}
    Because of \cref{ultimately-uniformly-recurrent:lem}, it suffices
    to decide whether $w$ is \kl{ultimately uniformly recurrent}.
    We can rewrite this as a question on the \kl{automatic sequence} $w$
    as follows:
    \begin{align*}
        &\exists N_0,                   &   \text{ultimately} \\
        &\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
        &\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
        &\exists k \geq 1,              &   \text{there exists a bound} \\
        &\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
        &\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
        &\exists l \geq 0,              &   \text{there exists a position in } v \\
        &\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
        &j_s + m + l < j_e \land
        w(i_s + m) = w(j_s + m + l) \quad .
    \end{align*}
    Because $w$ is computable by a finite automaton, one can reduce the above
    formula to a regular language, for which it suffices to check emptiness, which
    is decidable.
\end{proof}

One can go beyond \cref{automatic-wqo:thm} and consider
\kl(morphic){morphic sequences}, which are obtained by iterating a
\kl(substitution){substitution} on a finite alphabet, and then applying a
\kl(homomorphism){homomorphism}. The \kl{Fibonacci sequence} is an example of
a \kl(morphic){morphic sequence}. It is known that one can decide whether a
\kl(morphic){morphic sequence} is \kl{uniformly recurrent}
\cite[Theorem 1]{DURAND13}.

\todo[inline]{Define properly morphic words and 
  the questions associated to them. Also, there is a difference between
  \kl{ultimately uniformly recurrent} and \kl{uniformly recurrent}
  that is not handled here.}
\begin{theorem}
    \label{morphic-wqo:thm}
    Given a \kl(morphic){morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}.
\end{theorem}

>>>>

Let us first remark that if $S$ is a \kl{chain} for the \kl{suffix relation}
and $P$ is a \kl{chain} for the \kl{prefix relation}, then $SP$ is
\kl{well-quasi-ordered} for the \kl{infix relation}. This proves the (easy)
right-to-left implication of \cref{bounded-language:thm}.

@@

    Now, $\oWidth{L} \leq \sum_{i = 1}^n \oWidth{L_i} < 3 \cdot \omegaOrd <
    \omegaOrd^2$. We conclude that the \kl{ordinal width} of $L$ is less than
    $\omegaOrd^2$. Finally, the inequality $\oType{L} \leq \oWidth{L} \oComProd
    \oHeight{L} < \omega \oComProd \omega^2 = \omegaOrd^3$ allows us to conclude.

    The tightness of the bounds is a direct consequence of
    \cref{from-bi-to-single:lem}, and by considering a finite union of 
    these examples over disjoint alphabets (or even, by considering a binary 
    alphabet and using unambiguous codes to separate the different components).
\end{proofof}

\subsection{Decision Procedures}
\label{decision-procedures:sec}

\AP As we have demonstrated, infinite (or bi-infinite words) can be used to
represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation} by considering their set of finite \kl{infixes}.
Let us formalise the representation of languages by sets of bi-infinite
words that we will use in this section, following the characterization of
\cref{bi-infinite:lem}. A \intro{sequence representation} of a language $L
\subseteq \Sigma^*$ is a finite set of triples $(w_i^-, a_i, w_i^+)_{1 \leq i
\leq n}$ where $w_i^-, w_i^+ \in \Sigma^\Nat \cup \Sigma^*$ are two potentially
infinite words, and $a_i \in \Sigma$ is a letter, such that
\begin{equation*}
    L = \bigcup_{i = 1}^n \infset{\mathsf{reversed}(w_i^-) a_i w_i^+} \quad .
\end{equation*}

\AP Given an effective representation of sequences, one obtains an effective
representation of languages via \kl{sequence representations}. We can leverage
the theory of \kl{automatic sequences}, that is tailored to study the
regularity of infinite words to build a decision procedure for
\kl{well-quasi-ordered} languages for the \kl{infix relation}. A sequence $w
\in \Sigma^\Nat$ is \intro(sequence){automatic} whenever there exists a base $b
\in \Nat$ and a finite automaton that can compute the $i$-th letter of $w$
given as input the number $i$ in base $b$. The \kl{Thue-Morse sequence} is an
example of an \kl{automatic sequence}.

\begin{lemma}
  \label{automatic-uur:lem}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}
\begin{proof}
    We can rewrite this as a question on the \kl{automatic sequence} $w$
    as follows:
    \begin{align*}
        &\exists N_0,                   &   \text{ultimately} \\
        &\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
        &\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
        &\exists k \geq 1,              &   \text{there exists a bound} \\
        &\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
        &\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
        &\exists l \geq 0,              &   \text{there exists a position in } v \\
        &\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
        &j_s + m + l < j_e \land
        w(i_s + m) = w(j_s + m + l) \quad .
    \end{align*}
    Because $w$ is computable by a finite automaton, one can reduce the above
    formula to a regular language, for which it suffices to check emptiness, which
    is decidable.
\end{proof}

Based on \cref{automatic-uur:lem}, we can now prove the following theorem.
\begin{theorem}
    \label{automatic-wqo:thm}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{automatic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{theorem}
\begin{proof}
    It is easy to see that $L$ is \kl{well-quasi-ordered} for the \kl{infix
    relation} if and only if for every triple $(w_i^-, a_i, w_i^+)$ in the
    \kl{sequence representation} of $L$, the (potentially bi-infinite) word
    $\mathsf{reversed}(w_i^-) a_i w_i^+$ defines a \kl{well-quasi-ordered}
    language. By \cref{from-bi-to-single:lem}, this is the case if and only if
    both $w_i^-$ and $w_i^+$ are \kl{ultimately uniformly recurrent}. 
    Since one can decide whether an \kl{automatic sequence} is
    \kl{ultimately uniformly recurrent} from \cref{automatic-uur:lem},
    we conclude the proof.
\end{proof}

\todo[inline]{properly introduce morphic sequences, and maybe talk about
them in the introduction}

\AP
For \kl{automatic sequences}, the question of being \kl{ultimately uniformly
recurrent} is easily seen to be decidable, but symbolic dynamics provides a
more general framework to study infinite words. A \intro{morphic sequence} is
obtained by iterating a \kl(substitution){substitution} on a finite alphabet,
and then applying a \kl(homomorphism){homomorphism}. The \kl{Fibonacci
sequence} is an example of a \kl(morphic){morphic sequence}.
It was recently shown that one can decide whether a \kl(morphic){morphic
sequence} is \kl{uniformly recurrent} \cite[Theorem 1]{DURAND13}.
We can leverage this result to obtain the following lemma.

\begin{lemma}
  \label{morphic-uur:lem}
    Given a \kl(morphic){morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}

From \cref{morphic-uur:lem}, we can now prove the following corollary,
which is a generalization of \cref{automatic-wqo:thm}.

\begin{corollary}
    \label{morphic-wqo:cor}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl(morphic){morphic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}

### 2025-09-30 src/introduction.tex
new: morphic sequences factor complexity + moving stuff around

We then turn our attention to the \kl{infix relation}. In this case, we notice
that Lemma 5.1 from \cite{DBLP:journals/ita/Kuske06} imply that there are
\kl{well-quasi-ordered} languages for the \kl{infix relation} that have
arbitrarily large \kl{ordinal invariants} (except for the \kl{ordinal height},
which is always at most $\omegaOrd$). Therefore, we focus on two natural
semantic restrictions on languages: on the one hand, we consider \kl{bounded
languages}, that is, languages included in some $w_1^* \cdots w_k^*$ for some
finite choice of words $w_1, \ldots, w_k$; on the other hand, we consider
\kl{downwards closed} languages, that is, languages closed under taking
\kl{infixes}. In both cases, we provide a very precise characterization of
\kl{well-quasi-ordered} languages by the \kl{infix relation}, and derive tight
bounds on their \kl{ordinal invariants}. These results are summarized in
\cref{infixes-summary:fig}. We furthermore notice that for \kl{downwards
closed} languages that are \kl{well-quasi-ordered} by the \kl{infix relation},
being \kl(language){bounded} is the same as being \kl(language){regular}
(\cref{bounded-wqo-dwclosed:cor}), and that a \kl{bounded language} is
\kl{well-quasi-ordered} by the \kl{infix relation} if and only if its
\kl{downwards closure} is \kl{well-quasi-ordered} by the \kl{infix relation}
(\cref{bounded-wqo-dwclosed:cor}). This shows that, for \kl{bounded languages},
being \kl{well-quasi-ordered} implies that their \kl{downwards closure} is a
\kl{regular language}, which is a weakening of the usual result that the
\kl{downwards closure} of \emph{any language} for the \kl{scattered subword
relation} is always a \kl{regular language}.

@@

Let us point out that \cref{small-ordinal-invariants:thm} connects
\kl{well-quasi-orderings} by the \kl{infix relation} with the notion of
\kl{uniformly recurrent words}, borrowed from the field of word combinatorics.

Turning our attention to decision procedures, we consider two computational
models respectively tailored to \kl{downwards closed} languages and to
\kl{bounded languages}. For \kl{downwards closed} languages, we consider a
model based on \kl{morphic words}, for which we provide a decision procedure
(\cref{automatic-wqo:thm}). For \kl{bounded languages}, we consider the model
of \kl{amalgamation systems} \cite{ASZZ24}, which is an abstract computational
model that encompasses many classical ones, such as finite
automata, context-free grammars, and Petri nets \cite{ASZZ24}. We show that if
a language recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered}
by the \kl{infix relation}, then it is a \kl{bounded language}
(\cref{infix-amalgamation:thm}), and is therefore \kl{regular}. Furthermore, we
show that we can decide whether a given language recognized by an
\kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{infix relation}
(\cref{infix-wqo-is-emptiness:thm}). We defer the introduction of
\kl{amalgamation systems} to \cref{amalgamation-systems:subsec}.

>>>>

We then turn our attention to the \kl{infix relation}. In this case, we notice
that Lemma 5.1 from \cite{DBLP:journals/ita/Kuske06} imply that there are
\kl{well-quasi-ordered} languages for the \kl{infix relation} that have
arbitrarily large \kl{ordinal invariants} (except for the \kl{ordinal height},
which is always at most $\omegaOrd$). Therefore, we focus on two natural
semantic restrictions on languages: on the one hand, we consider \kl{bounded
languages}, that is, languages included in some $w_1^* \cdots w_k^*$ for some
finite choice of words $w_1, \ldots, w_k$; on the other hand, we consider
\kl{downwards closed} languages, that is, languages closed under taking
\kl{infixes}. In both cases, we provide a very precise characterization of
\kl{well-quasi-ordered} languages by the \kl{infix relation}, and derive tight
bounds on their \kl{ordinal invariants}. These results are summarized in
\cref{infixes-summary:fig}. \todo{this is not what the corollary says} We furthermore notice that for \kl{downwards
closed} languages that are \kl{well-quasi-ordered} by the \kl{infix relation},
being \kl(language){bounded} is the same as being \kl(language){regular}
(\cref{bounded-wqo-dwclosed:cor}), and that a \kl{bounded language} is
\kl{well-quasi-ordered} by the \kl{infix relation} if and only if its
\kl{downwards closure} is \kl{well-quasi-ordered} by the \kl{infix relation}
(\cref{bounded-wqo-dwclosed:cor}). This shows that, for \kl{bounded languages},
being \kl{well-quasi-ordered} implies that their \kl{downwards closure} is a
\kl{regular language}, which is a weakening of the usual result that the
\kl{downwards closure} of \emph{any language} for the \kl{scattered subword
relation} is always a \kl{regular language}.

@@

Turning our attention to decision procedures, we consider two computational
models respectively tailored to \kl{downwards closed} languages and to
\kl{bounded languages}. For \kl{downwards closed} languages, we consider a
model based on representations of infinite words
(\cref{decision-procedures:sec}), for which we provide a decision procedure
(\cref{automatic-wqo:thm}). The model used to represent these infinite words is
based on \kl{automatic sequences} and \kl{morphic sequences} \cite{CAKA97},
which are well-studied in the context of symbolic dynamics.
For \kl{bounded languages}, we consider the model
of \kl{amalgamation systems} \cite{ASZZ24}, which is an abstract computational
model that encompasses many classical ones, such as finite automata,
context-free grammars, and Petri nets \cite{ASZZ24}. We show that if a language
recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the
\kl{infix relation}, then it is a \kl{bounded language}
(\cref{infix-amalgamation:thm}), and is therefore \kl{regular}. Furthermore, we
show that we can decide whether a given language recognized by an
\kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{infix relation}
(\cref{infix-wqo-is-emptiness:thm}). We defer the introduction of
\kl{amalgamation systems} to \cref{amalgamation-systems:subsec}.
